---
title: "Does three point shooting determine NCAA Tournament game outcomes more than any other factor?"
author: "Jon Michael Stroh"
date: "`r Sys.Date()`"
output: 
 pdf_document: 
    fig_height: 3
    fig_width: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = "center")
```

```{r}
library(tidyverse)
library(broom)
library(knitr)
library(readxl)
library(patchwork)
library(Stat2Data)
library(yardstick)
```

```{r}
marchmadness2021 <- read_excel("Data/March_Madness_2021_AllGames_Data.xlsx")
mm2021byteams <- 
read_excel("Data/March_Madness_2021_GamesByTeam_Data.xlsx")
mm2021byteams <- mm2021byteams %>% mutate(Win = factor(Win))
```
## Data Exploration (2021 NCAA Tournament)

### New Variables

#### More3s
Frequency of the winning team out shooting the losing team on 3 point shots (variable more3s equals 0 if losing team made more three point shots, 0.5 if an equal amount were made, and 1 if the winning team made more three point shots):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(more3s = factor(if_else(W3P > L3P, 1, if_else(W3P == L3P, 0.5, 0))))

marchmadness2021 %>% count(more3s)
```

In the 2021 tournament, the winning team made more three pointers in 63.6% (42/66) of games, the teams tied in 13.6% (9/66) of games, and the losing team made more three pointers in 22.7% (15/66) of games.

#### Better3s

Frequency of the winning team out shooting the losing team on 3 point shot percentage (variable better3s equals 0 if the losing team made a higher percentage of three point shots, 0.5 if an equal percentage were made, and 1 if the winning team made a higher percentage):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(better3s = factor(if_else( (W3P/W3PA) > (L3P/L3PA), 
  1, if_else((W3P/W3PA) == (L3P/L3PA), 0.5, 0))))

marchmadness2021 %>% count(better3s)
```

In the 2021 tournament, the winning team made a higher percentage of three pointers in 77.3% (51/66) of games, while the losing team made a higher percentage in only 22.7% (15/66) of games.

#### More & Better 3s

```{r}
marchmadness2021 %>% count(more3s, better3s, sort = TRUE)
```

In 2021, in only 13.6% (9/66) of games did the winning team make less 3s and a lower percentage of them.  In contrast, in 68.2% (45/66) of games did the winning team either make more 3s and shoot a higher percentage (38 games) or tie on the number of makes and shoot a higher percentage (7 games).

#### More 2s
Frequency of winning team out shooting losing team on 2 point shots (variable more2s equals 0 if losing team made more two point shots, 0.5 if equal amount were made, and 1 if the winning team made more two point shots):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(more2s = factor(if_else(W2P > L2P , 1, if_else(W2P == L2P, 0.5, 0))))

marchmadness2021 %>% count(more2s)
```

In only 42 games did the winning team make more 2 point baskets, and in 20 games the losing team made more (4 ties).

#### Better 2s
Frequency of winning team out shooting losing team on 2 point shot percentage (variable better2s equals 0 if losing team made a higher percentage of two point shots, 0.5 if equal amount were made, and 1 if the winning team a higher percentage of two point shots):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(better2s = factor(if_else((W2P/W2PA) > (L2P/L2PA), 1, if_else((W2P/W2PA) == (L2P/L2PA), 0.5, 0))))

marchmadness2021 %>% count(better2s)
```

WOW-- in only 44 games did the winning team out shoot the losing team on 2 point percentage, and in 21 games, the losing team out shot the winning team on 2 point percentage.

#### moreFgs
Frequency of the winning team out shooting the losing team in general (more shots made) (variable morefgs equals 0 if losing team made more shots, 0.5 if equal amount were made, and 1 if the winning team made more shots):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(morefgs = factor(if_else(WFG > LFG , 1, if_else(WFG == LFG, 0.5, 0))))

marchmadness2021 %>% count(morefgs)
```

In 84.8% of games the winning team made more field goals (makes sense, only other option to win then is by making more 3s as a portion of those field goals or more free throws or both).  There were 3 ties.

#### betterfgs
The frequency of the winning team out shooting the losing team by percentage (higher percentage of total shots made) (variable betterfgs equals 0 if losing team made more shots, 0.5 if equal amount were made, and 1 if the winning team made more shots):

```{r}
marchmadness2021 <- marchmadness2021 %>% 
  mutate(betterfgs = factor(if_else((WFG/WFGA) > (LFG/LFGA), 
  1, if_else((WFG/WFGA) == (LFG/LFGA), 0.5, 0))))

marchmadness2021 %>% count(betterfgs)
```

However, interesting enough, in only 71.2% of games (47/66) did the winning team shoot a higher percentage on all field goals. With this being the case, clearly other factors can influence a win, the likely ones for investigation is free throw shooting (although there tends to be a small difference in this between teams in a single game), three point shooting (our variable of primary interests), and factors that can get a team more shots (offensive rebounding and forcing turnovers while limiting those things for themselves).



### Distribution of Shooting

#### Two point shooting
twoPercent is the two point shooting percentage by team as a percentage, rather than a decimal, for graphing.

```{r}
mm2021byteams <- mm2021byteams %>% mutate(twoPercent = `2P%` * 100)

ggplot(data = mm2021byteams, aes(x = twoPercent)) + 
  geom_histogram(bins = 20) + 
  labs(x = "Shooting Percentage", 
       title = "Two Point Shooting Percentages in MM 2021")
```

#### Three point shooting
threePercent is the three point shooting percentage by team as a percentage, rather than a decimal, for graphing.

```{r}
mm2021byteams <- mm2021byteams %>% mutate(threePercent = `3P%` * 100)

ggplot(data = mm2021byteams, aes(x = threePercent)) + 
  geom_histogram(bins = 20) + 
  labs(x = "Shooting Percentage", 
       title = "Three Point Shooting Percentages in MM 2021")
```

#### Comparing Shooting

*Note: twoPercent_mean is not the shooting percent for all two point shots in the march madness tournament.  It is the mean shooting two point shooting percentage for each team each game.  Overall, it can not represent the percentage for total two point shooting percent because it is not adjusted for attempts by game.  The same thing goes for threePercent_mean*

```{r}
mm2021byteams %>%
  summarise(twoPercent_Mean = mean(twoPercent),
            threePercent_Mean = mean(threePercent),
            twoPercent_SD = sd(twoPercent),
            threePercent_SD = sd(threePercent))
```

Although three point shooting appears to much more closely resemble the normal distribution, it has a larger standard deviation indicating that there is more variation in three point shooting than two point shooting (although the standard deviations are very close) by game.  According to the standard deviations, 68% of the team's three point shooting by game occurs within 10.97% of 33.45%, while 68% of the team's two point shooting by game occurs within 10.14% of 48.97%.  There is slightly more variation is three point shooting performances by game than two point shooting performances.

#### Three Point Percentage by Outcome

```{r}
ggplot(data = mm2021byteams, aes(y = `3P%`, color = Win)) + 
  geom_boxplot() + 
  labs(y = "Shooting Percentage", 
       title = "Three Point Shooting Percentages by Outcome in MM 2021")
```

```{r}
with(mm2021byteams, aggregate(`3P%` ~ Win, FUN = median))
```

The median three point percentage for winning teams is 37.2% compared to only 28.6% for losing teams.

Based on data exploration for all 66 games in the 2021 NCAA tournament, there seems to be a strong association between a team's three point shooting percentage for a single game and the probability that they win that game.  Thus, we intend to explore models with three point percentage as a predictor variable for whether a team wins an NCAA tournament game or not.

## Model Building

According to a 2004 study, 4 factors were indicated to most affect the outcome of basketball games: "shooting efficiency, number of turnovers, offensive rebounds and free throws made" (Oliver).  Thus, I will fit a logistic regression model that includes two and three point percentage, total turnovers, total offensive rebounds, and total free throws attempted as predictor variables for whether or not a team wins a game.  Additionally, through my own observation of college basketball, assisted shots tend to be better shots, and thus more likely to go in. Therefore, I will add total assists to my model as a predictor.  Finally, teams that generate many steals tend to generate easy, fast break scoring opportunities.  Also, teams that limit fouls often limit the amount of times a team goes to the free throw line.  Thus, I will add total steals and total fouls to the model.  Finally, I want to mean center all the variables in the model so that the intercept can be interpreted.


```{r}
mm2021byteams <- mm2021byteams %>% 
  mutate(twoPercentCent = twoPercent - mean(twoPercent), threePercentCent = threePercent - mean(threePercent), ORBCent = ORB - mean(ORB), ASTCent = AST - mean(AST), TOVCent = TOV - mean(TOV), STLCent = STL - mean(STL), PFCent = PF - mean(PF), FTACent = FTA - mean(FTA))
```

```{r}
full_mm2021_model <- glm(Win ~ twoPercentCent + threePercentCent + ORBCent + ASTCent + TOVCent + STLCent + PFCent + FTACent, data = mm2021byteams, family = binomial)

tidy(full_mm2021_model) %>% kable(digits = 3)
```

According to the model, the predictor variables with statistical significance for predicting whether a team won or loss is two point percentage, three point percentage, total turnovers, total steals, total fouls, and free throws attempted.

According to the model, the odds of a team winning the game with the mean predictor values -- `r mean(mm2021byteams$twoPercent)`% two point shooting, `r mean(mm2021byteams$threePercent)`% three point shooting, `r mean(mm2021byteams$ORB)` offensive rebounds, `r mean(mm2021byteams$AST)` assists, `r mean(mm2021byteams$TOV)` turnovers, `r mean(mm2021byteams$STL)` steals, and `r mean(mm2021byteams$PF)` fouls -- is `r round(exp(0.014), 3)`.  This value makes sense because a team with the mean statistics in this value has an almost 1 to 1 odds of winning a game or roughly a 50% win percentage.

According to the model, for every 1 percentage point increase in two point shooting percentage the odds the team wins the game is multiplied by a factor of `r round(exp(0.110), 3)`, holding all else constant.  Similarly, according to the model, for every 1 percentage point increase in three point shooting percentage the odds the team wins the game is multiplied by a factor of `r round(exp(0.100), 3)`, holding all else constant.  Thus, the coefficients for both two and three point shooting percentages are close, but the factor in which the odds of winning are multiplied by are slightly larger for two point shooting than three.

Now, to fit a better model, I will conduct backwards model selection using AIC as well as a drop in deviance test to attempt to find the strongest model:

#### Backwards Selection

```{r}
selected_mm2021_model_AIC <- step(full_mm2021_model, direction = "backward", trace = FALSE)
tidy(selected_mm2021_model_AIC) %>% kable(digits = 3)
```
#### Drop-in-deviance test

Hypotheses for Drop-in-deviance test:

Null Hypothesis: $H_0: \beta_{ORBCent} = \beta_{ASTCent} = 0$
(These variables don't add information to the model after accounting for two & three point shooting percentage, steals, turnovers, fouls, and free throws attempted)

Alternative Hypothesis: $H_a: \text { at least one } \beta_j \text { is not equat to 0}$

```{r}
tidy(anova(selected_mm2021_model_AIC, full_mm2021_model, test = "Chisq"))
```

Because the p-value is much larger than 0.05, we fail to reject the null hypothesis.  That data does not provide sufficient evidence that the  coefficients for ORBCent and ASTCent are different from 0. The best model includes only two & three point shooting, turnovers, steals, fouls, and free throw attempts.

#### Model Interpretation

```{r}
tidy(selected_mm2021_model_AIC) %>% kable(digits = 3)
```

Again, according to the model, the factor in which the odds of winning are multiplied by for each 1 percentage point increase in shooting percentages of twos and threes is still greater for an increase in two point percentage (`r round(exp(0.095), 3)` versus `r round(exp(0.090), 3)`).  Although, the coefficients are very close.

To make predictions for team's game statics, we can refit the model without mean centering the variables.

```{r}
selected_mm2021_model_AIC_nocenter <- glm(Win ~ twoPercent + threePercent + TOV + STL + PF + FTA, data = mm2021byteams, family = binomial)
tidy(selected_mm2021_model_AIC_nocenter) %>% kable(digits = 3)
```

For example, in the first round of the NCAA tournament, 1 seed Baylor beat 16 seed Hartford 79 to 55.  According to the model, the odds that Baylor would win the game given its game statistics are expected to be  `r round(exp(1.777), 3)` ($exp^{-5.473 + 0.095(47.6) + 0.090(33.3) - 0.214(10) + 0.277(15) - 0.229(16) + 0.138(10)}$).  This indicates that given its game statistics, Baylor is expected to win `r round(exp(1.777), 3)` games for every one loss. Similarly, given the same formula, according to the model, the odds that Hartford would win the game given its game statistics are expected to be `r round(exp(-4.38975), 3)`.  Thus, Hartford would be expected to win `r round(exp(-4.38975), 3)` games for every loss, or in a better interpretation, 1 win the team is expected to lose roughly 83.3 games.  This may seem extreme but Hartford did have a very poor performance (24 turnovers!), and so it is unlikely a team will win many games with 24 turnovers and such poor shooting.

### Model Conditions

#### Linearity

To check if linearity is satisfied, we plot the predictor variable against the empirical logit and are looking for a linear relationship.

```{r}
emplogitplot1(Win ~ twoPercent, data = mm2021byteams, ngroups = 10)
```
```{r}
emplogitplot1(Win ~ threePercent, data = mm2021byteams, ngroups = 10)
```
```{r}
emplogitplot1(Win ~ TOV, data = mm2021byteams, ngroups = 10)
```
```{r}
emplogitplot1(Win ~ FTA, data = mm2021byteams, ngroups = 10)
```
```{r}
emplogitplot1(Win ~ PF, data = mm2021byteams, ngroups = 5)
```
```{r}
emplogitplot1(Win ~ STL, data = mm2021byteams, ngroups = 5)
```
Based on the empirical logit plots, the linearity condition is met for each predictor variable.

#### Randomness

The randomness condition is murky. Each individual basketball is game is a random event.  However, teams are inherently unequal, and therefore each game is not a 50-50 random event.

#### Indepedence

Again, murky.  The outcome of one game does not directly affect the outcome of another game.  However, it does effect the teams that play, with better teams typically advancing.

### Assessing Model Fit

```{r}
mm2021byteams_aug <- augment(selected_mm2021_model_AIC_nocenter, type.predict = "response")

mm2021byteams_aug <- mm2021byteams_aug %>% mutate(pred_Win = if_else (.fitted > 0.55, "Predicted Win", "Predicted Loss"))

mm2021byteams_aug %>% count (Win, pred_Win)
```

According to the confusion matrix for the model with a threshold of 0.55, the misclassification rate is 17.4%, with a sensitivity of 81.8% and a specificity of 83.3%.  The threshold 0.55 was observed as the best to minimize misclassification as well as relatively equal sensativity and specificity.

```{r}
mm2021byteams_aug <- mm2021byteams_aug %>% mutate(Win = factor(Win, levels = c("1", "0")))

roc_curve_mm2021byteams_aug <- mm2021byteams_aug %>%
  roc_curve(Win, .fitted) 

autoplot(roc_curve_mm2021byteams_aug)
```
```{r}
mm2021byteams_aug %>%
roc_auc(Win, .fitted) %>%
 pull(.estimate)
```

According to the ROC Curve and the AUC of 0.887, the model is a good fit for the data.  Considering there are only six predictor variables (the model being fairly simple), AUC values closer to 1 and farther from 0.5 indicate better model fit for the data.

#### Benefits and Drawbacks to this Model

Certain benefits to this model include: its applicability to predicting games in the NCAA tournament.  To predict future NCAA tournament games, one can apply the average statistics for a team to predict whether or not they will win the game.  Additionally, its accuracy is based upon the previous NCAA tournament and what current trends/developments can be more applicable than the previous tournament.  Furthermore, the model is very simple.

However, the drawbacks:  first, the 20% error rate.  But, more importantly, the quantitative statistcs other than shooting percentages -- turnovers, steals, fouls, and free throw attempted -- require pace adjusting that the model can not take into account.  Teams that play at a faster pace will have a greater total number of raw statistics-- shots, rebounds, and yes, turnovers, steals, fouls, and free throw attempted.  Thus, it is important if making predictions to enter per 40 minute totals to the model to make predictions.

What this model does not successfully do, which was the purpose of this study, is indicate whether three point shooting has a greater affect than any other factor in determining the outcome of NCAA tournament games.  However, assigning value to different factors in the importance for wins is difficult. This model did indicate that of the four factors, shooting percentages, offensive rebounds, turnovers, and free throws attempted,  offensive rebounds was not statistically significant and left out of the model. 










